An Answer Set Programming Framework for Reasoning about Agents' Beliefs and Truthfulness of Statements

### 2020-03-12 main.tex
Updates from Overleaf

%
\documentclass{article}
\usepackage[utf8]{inputenc}

%>>>>>>>>>> AAAI format

\usepackage{aaai20}
\usepackage{times}
\usepackage{helvet}
\usepackage{courier}
\usepackage[hyphens]{url}
\usepackage{graphicx}
\urlstyle{rm}
\def\UrlFont{\rm}
\usepackage{graphicx}
\frenchspacing
\setlength{\pdfpagewidth}{8.5in}
\setlength{\pdfpageheight}{11in}
% Add additional packages here, but check
% the list of disallowed packages
% (including, but not limited to
% authblk, caption, CJK, float, fullpage, geometry,
% hyperref, layout, nameref, natbib, savetrees,
% setspace, titlesec, tocbibind, ulem)
% and illegal commands provided in the
% common formatting errors document
% included in the Author Kit before doing so.
%
% PDFINFO
% You are required to complete the following
% for pass-through to the PDF.
% No LaTeX commands of any kind may be
% entered. The parentheses and spaces
% are an integral part of the
% pdfinfo script and must not be removed.
%
%\pdfinfo{
%/Title (Type Your Paper Title Here in Mixed Case)
%/Author (John Doe, Jane Doe)
%/Keywords (Input your keywords in this optional area)
%}
%
% Section Numbers
% Uncomment if you want to use section numbers
% and change the 0 to a 1 or 2
% \setcounter{secnumdepth}{0}

% Title and Author Information Must Immediately Follow
% the pdfinfo within the preamble
%
\title{An Answer Set Programming Framework for Reasoning about Truthfulness of Statements by Agents} %\\
\author{Marcello Balduccini\\
Saint Joseph's University\\
\And
Michael Gelfond\\
Texas Tech University\\
\And
Enrico Pontelli \ and Tran Cao Son\\
New Mexico State University
} %\\

@@

%\usepackage[nodisplayskipstretch]{setspace}
%\setstretch{1}

\setlength{\belowdisplayskip}{0pt} 
\setlength{\belowdisplayshortskip}{0pt}
\setlength{\abovedisplayskip}{0pt} 
\setlength{\abovedisplayshortskip}{0pt}

@@

\setlength{\dbltextfloatsep}{-1pt}  
\setlength{\dblfloatsep}{-1pt}

@@

@@

@@

Artificial Intelligence research, in the area of knowledge-based and intelligent agents, has been progressing at
a rapid pace, making it possible to develop agents that can assist and/or replace humans in 
several tasks. Organizations  use  web-bots for interacting with clients in various capacities, e.g., in providing information about the company or making offers. This trend continues to be fueled by the ubiquitous use of online resources. 
Unfortunately, not every business on the Internet is as 
honest as one would hope. Stories about businesses that cheat people of goods or services, or falsely advertise their services are not uncommon. Indeed, lying and misrepresentations are more widespread in online negotiations than 
offline ones \cite{elie}. This concern is also reflected in the 
development of resources that allow people to rate companies (e.g., \url{expedia.com}, \url{yelp.com},  \emph{Angie's List}) or defend the reputation of a company or an entity (e.g., \url{reputation.com}). Dishonesty is also present in
the news media, leading to the dissemination of untruthful stories  (e.g., the `Phuc Dat Bich' story\footnote{ 
{\tiny
\url{http://gizmodo.com/phuc-dat-bich-is-a-massive-phucking-faker-1744588099}
}} reported by BBC and Sky News). Another example is the case of a hotel reviewer with the alias {\em ``AmishBoy''} who gave high ratings to different hotels in the same day \cite{MinnichCMLF15}---thus suggesting  that his statements may be untrustworthy. These situations highlight the problem of  \emph{understanding whether 
statements made by an agent can be accepted as truthful wth respect to a context composed of observations,
knowledge about default behavior, and corresponding beliefs. }

@@

\begin{example} \label{ex1}
When we first met at an event (time $t_0$), John stated that he comes from a poor family. 
We later learn that John attends a private college, renowned for  its expensive tuition (time $t_1$). Later on, we learn 
that John attends the college thanks to a  scholarship awarded  due to his financial hardship (time $t_2$).

This story spans over three time instances: $t_0$, $t_1$, and $t_2$. During the story, we learn (through observations) some facts, which  affect our beliefs about the world as follows:

\begin{list}{$\bullet$}{\itemsep=0pt \topsep=0pt \parsep=0pt \leftmargin=10pt}  
\item Time $t_0$: John makes the statement that his family is poor (property $poor$ is true). No observations are available to us and thus we do not know if John's family is indeed poor or not. We have no reasons to conclude that John's statement is not truthful.

\item Time $t_1$:  We observe that John attends an expensive college (property $in\_college$ is true). {\emph{Normally}, we believe that someone attending an expensive college is from a rich family.} (default $d_1$). Hence, our belief at this point is that John is rich. This prevents us from accepting John's statement as truthful. We indicate that  the default $d_1$ is the reason to draw such conclusion.

\item Time $t_2$: We observe that John has a need-based scholarship   (property $has\_scholarship$ is true). {It is our belief that a student's need-based scholarship is \emph{usually} derived from the family's financial situation} (default $d_2$). Because both defaults are applicable and in conflict with each other, we neither believe that John's family is rich, nor that it is poor. This allows us to withdraw our previous conclusion that John's statement could not be accepted as truthful.

\item The situation might be different if we had a preference between our defaults. For instance, if we were inclined to favor $d_2$ over $d_1$, then we would believe that John's family is poor. This may strengthen our conclusion that John's statement can be accepted as truthful. 
% 
\end{list} 
\end{example}  
The main contributions of this paper can be summarized as follows:
\begin{enumerate}
\item The formalization of an abstract model to 
represent and reason about the evolution of an agent's beliefs over time and to draw conclusions about whether third-party statements can be accepted as truthful; 
\item A concrete realization of the model using
Answer Set Programming; 
\item A use case of the proposed framework on an important problem in the
        area of  cyber security. 
\end{enumerate}
%The paper illustrates  how, starting
% from ({\em (i)}) a collection of observations about the (possibly evolving) state of the
% world, {\em (ii)}
% observations about (some of) the  actions performed by the agent, and 
%{\em (iii)}
% the observer's own bias about the \emph{normal} behavior of agents,
% an observer can draw conclusions about the truthfulness of statements made by
% the observed agent. 
 We illustrate the framework 
 using examples and discuss possible extensions that need to be considered. 
%We also discuss a potential application of the proposed framework.

@@

The language of $\cal B$ is based on a signature ${\cal K} = ({\cal L}, {\cal A})$, where
$\cal L$ is a set of ground atoms, referred to as \emph{fluents},
 and $\cal A$ is a set of \emph{action names}. A fluent literal is a literal
 built on $\cal L$.
 A domain description of $\cal L$ is composed of a collection of \emph{axioms} of 
 the following types:
 \begin{list}{$\bullet$}{\topsep=1pt \itemsep=1pt \parsep=0pt \leftmargin=10pt}
 \item \emph{Executability Axioms:} these describe the conditions that enable the
         execution of an action, and they have the form $a \executable \ell_1\wedge \dots\wedge \ell_n$,
         where $a\in {\cal A}$ and $\ell_1,\dots,\ell_n$ are fluent literals.
  \item \emph{Dynamic Causal Laws:} these descriptions the direct effects of the execution
          of an action, and they have the form
          $a \causesif{\ell} \ell_1\wedge \dots \wedge \ell_n$
          where $a\in {\cal A}$ and $\ell,\ell_1,\dots,\ell_n$ are fluent literals.
  \item \emph{Static Causal Laws:} these description static dependencies among fluents, and they
          have the form
          $\ell \ifthen \ell_1\wedge\dots\wedge\ell_n$
          where $\ell, \ell_1,\dots,\ell_n$ are fluent literals.
 \end{list}
 A planning problem in $\cal K$ is a tuple $(D, I, O)$, where $D$ is a domain description
 of $\cal K$ and $I, O$ are two consistent sets of fluent literals---denoting, respectively,
 the initial state and the goals of the planning problem.

@@

In this section, we propose a general framework for representing, and reasoning about, the beliefs held by an agent. The framework also establishes a link between the beliefs held by the agent and whether it can accept as truthful statements made by a third party. The
framework  can be instantiated using specific paradigms for reasoning about actions and change and for non-monotonic reasoning. We make
the following assumptions:
%
 \begin{itemize}
%
\item It is possible to  observe the properties of the world and the occurrences of the actions    over time
        (e.g., we observe that John buys a car, John is a student, etc.). Let us denote with $O_a$ and $O_w$ 
        the set of action occurrences and the set of observations about the world over time, respectively.

\item We have adequate knowledge about actions  and their effects (e.g., the action of buying a car requires that the agent has money and its execution will result in the agent owning a car). This knowledge is represented by an action theory $Act$ in a suitable logic $A$ that allows  reasoning about actions' effects and consequent changes to the world. Let us denote with $\models_A$ the entailment relation defined within the logical framework $A$ used to describe $Act$.

@@

The set of observations $O_w$ in Example~\ref{ex1} includes the observations such as 
%`John comes from a poor family' at time point $t_0$, 
`John attends an expensive college' at time point $t_1$, and `John receives a need-based scholarship' at time point $t_2$. In this particular example we do not have any  action occurrences, i.e., $O_a = \emptyset$.\footnote{Some frameworks for reasoning about actions and change may require the introduction of a NOP action for modeling the example.} Our default theory $D$ consists of defaults $d_1$ and $d_2$ stated earlier, which allow us to make conclusions regarding our agent's beliefs.

%TEMP
\begin{figwindow}[1,r,%
        {\includegraphics[width=0.48\textwidth]{newstep.pdf}},%
        {Answering Question: $(O_a,O_w,Act,Def) \models p[t]$?\label{fig:schema} }]
\end{figwindow}
Let us consider a theory $T = (O_a, O_w, Act, \textit{Def})$ and the associated entailment relations $\models_A$ and $\models_D$. First and foremost, we are interested in defining the beliefs held by the agent over time. And, then, we are interested in using this information for answering the question of whether a statement $p[t]$,  made at time step $t$ about a proposition $p$, can be accepted as truthful. For this, we define the entailment relation $\models$ between $T$ and $p[t]$. Intuitively, the process entails the steps below (Figure~\ref{fig:schema}):
\begin{list}{$\bullet$}{\itemsep=0pt \parsep=0pt \topsep=0pt \leftmargin=8pt} 
\item Compute possible models $W[t]$ of the world  at the time step $t$ from  $Act$, $O_a$, and $O_w$ (using $\models_A$); and  
\item Determine  the agent's beliefs by finding which propositions are true  given  $\textit{Def}$ and $W[t]$ (using $\models_D$); \textbf{CHECK FOR ISSUES DUE TO THINGS THAT ARE *TRUE* FROM W[t]}
\item Determine if the statement $p[t]$ is consistent with respect to the agent's beliefs. 
\end{list} 
%
Let us assume that the equation $W[t] = \{z \mid Act \cup O_a \cup O_w \models_A z[t]\}$ characterizes
any of the states of the world  at  time step $t$ given $Act$, $O_a$, and $O_w$ (based on the semantics of 
$\models_A$). The entailment relation between $T$ and $p[t]$ can be defined as follows.
%
{\begin{equation} \label{entail} 
T \models p[t] \Leftrightarrow
\begin{array}{c} 
%T \models p[t] \\
% \textit{if and only if} \\
  \forall W[t]. \left(\begin{array}{c}
   W[t] = \{z \mid Act \cup O_a \cup O_w \models_A z[t]\} \Rightarrow \\
                                                                 Def \cup W[t] \models_D p
                                                                \end{array}\right)
\end{array} 
\end{equation} }
%
Note that this definition also allows one to identify elements of $O_a$ and $O_w$ which, when obtained, will result in the 
confirmation or denial of $T \models p[t]$. As such, a system that obeys \eqref{entail} can also be used by users who are 
interested in  what they need to do in order to believe in a statement about $p$ at the time step $t$,
 given their beliefs about the behavior of the observed agents.

%
%\smartdiagram[flow diagram:horizontal]{Edit,
%  \LaTeX, Bib\TeX/ biber, make\-index, \LaTeX}
\iffalse\begin{figure}
\centering
\resizebox{1.6in}{!}{% 
\smartdiagram[priority descriptive diagram]{
  $W[t]$ model of $A \cup O_a \cup O_w$ restricted to time step $t$,
  $\forall W[t]. (D \cup W[t] \models_D p)$?}
}  
\caption{Answering $(O_a,O_w,A,D) \models p[t]$?}
\label{fig:schema} 
\end{figure} \fi
%
To develop a concrete system for reasoning about truthfulness of agents using \eqref{entail}, specific formalizations of $Act$ and $\textit{Def}$ need to be developed. There is a large body of research related to these two areas, and deciding which one to use depends on the system developer. 
Well-known formalisms for reasoning about actions and change, such as action languages \cite{GelfondL98}, situation calculus \cite{Reiter01}, etc., can be employed for $Act$ (and $\models_A$).  
Approaches to default reasoning with preferences, such as those proposed in  \cite{brew99,BrewkaE00,DelgrandeST03,GelfondS98}), can be used for $\textit{Def}$ (and $\models_D$). In addition, 
let us note that, in the literature, $\models_D$ can represent \emph{skeptical} or \emph{credulous} reasoners; and  the model does not specify how observations are collected. Deciding which type of reasoning is suitable or how to collect observations is an important issue, but it is application-dependent and beyond  the scope of this paper. 
%In this paper, we use observations  in a broad sense: it might be something that we learn some facts about a property of the world, or we actually observed it (via the execution of an action), etc. 
%
\section{Reasoning about Truthfulness of Agents Using Answer Set Programming (ASP)}

We will now present a concrete system for reasoning about truthfulness of agents using \emph{Answer Set Programming (ASP)}.  We select ASP as the host language since there exist several approaches to reasoning about actions and changes as well as default reasoning using ASP. This enables  the seamless integration of the two steps involved in ~\eqref{entail} into a single system. 
Several efficient ASP solvers are available, that enable the use of the proposed system in practical applications. In particular, we will use answer sets (and projections of answer sets over specific time points) to capture $W[t]$, and use established encodings of action theories and defaults in ASP.

@@

ASP is a programming paradigm \cite{Niemela99,MarekT99}  based on logic programming under the answer set semantics \cite{GelfondL90}. 
A logic program $\Pi$ is a set of rules of the form 
 $$c  \leftarrow a_1,\ldots,a_m,\naf a_{m+1},\ldots,\naf a_n$$ 
%
where $0 \le m \le n$,   $a_i$'s and $c$ 
are  ground literals in a first order language,\footnote{Rules with variables represent the set of all 
        of their  ground instances.}   
and $\naf$ represents {\em default negation}. 
A rule is satisfied by a set of literals $S$ if $c \in S$, or $\{a_1,\ldots,a_m\} \setminus S \ne \emptyset$, or $\{a_{m+1},\ldots,a_{n}\} \cap S \ne \emptyset$. A consistent set of literals $S$ is a model of a program if it satisfies all the rules of the program. $S$ is an answer set of a program if it is a consistent minimal model of the program obtained by {\em (i)} deleting all rules with some $a \in S$ such that $not \: a$ appears in the rule; and {\em (ii)} deleting all  $not \: a$ from the remaining rules.  Programs without an answer set are said to be inconsistent. Answer sets can be computed using answer set solvers. Extended syntactic notations have been proposed to facilitate the encoding of commonly used patterns (e.g., aggregates). In this paper, we use the  syntax of ASP used in  {\sc Clasp}\footnote{
  \url{http://potassco.sourceforge.net/}
} or Dlv\footnote{\url{http://www.dlvsystem.com/}
}.

>>>>

%
%%%% kr.tex

\typeout{KR--20 Instructions for Authors}

% These are the instructions for authors for KR-20.

\documentclass{article}
\pdfpagewidth=8.5in
\pdfpageheight=11in

\usepackage{kr}

% Use the postscript times font!
\usepackage{times}
\usepackage{soul}
\usepackage{url}
\usepackage[hidelinks]{hyperref}
\usepackage[utf8]{inputenc}
\usepackage[small]{caption}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{booktabs}
%\usepackage{algorithm}
\usepackage{algorithmic}
%\usepackage{algorithmicx}
\urlstyle{same}

% Add additional packages here, but check
% the list of disallowed packages
% (including, but not limited to
% authblk, caption, CJK, float, fullpage, geometry,
% hyperref, layout, nameref, natbib, savetrees,
% setspace, titlesec, tocbibind, ulem)
% and illegal commands provided in the
% common formatting errors document
% included in the Author Kit before doing so.
%
% PDFINFO
% You are required to complete the following
% for pass-through to the PDF.
% No LaTeX commands of any kind may be
% entered. The parentheses and spaces
% are an integral part of the
% pdfinfo script and must not be removed.
%
%\pdfinfo{
%/Title (Type Your Paper Title Here in Mixed Case)
%/Author (John Doe, Jane Doe)
%/Keywords (Input your keywords in this optional area)
%}
%
% Section Numbers
% Uncomment if you want to use section numbers
% and change the 0 to a 1 or 2
% \setcounter{secnumdepth}{0}

@@

%\usepackage[nodisplayskipstretch]{setspace}
%\setstretch{1}
\iffalse 
\setlength{\belowdisplayskip}{0pt} 
\setlength{\belowdisplayshortskip}{0pt}
\setlength{\abovedisplayskip}{0pt} 
\setlength{\abovedisplayshortskip}{0pt}

@@

\setlength{\dbltextfloatsep}{-1pt}  
\setlength{\dblfloatsep}{-1pt}
\fi

@@

\allowdisplaybreaks

@@

\usepackage{listings}

\definecolor{darkgray}{rgb}{0.66, 0.66, 0.66} % needed for a clean compile
\lstdefinelanguage{clingo}{
  keywordstyle=[1]\usefont{OT1}{cmtt}{m}{n},%
  keywordstyle=[2]\textbf,%
  keywordstyle=[3]\usefont{OT1}{cmtt}{m}{n},%\textit
  alsoletter={\#,\&},%
  keywords=[1]{not,from,import,exists,if,else,return,while,
                break,and,or,for,in,del,and,class,subClass,concern,aspect,subCo,prop,rdf,cpsf,addBy,suppBy,neg,d,relation,holds,h,obs,action,fluent,occurs,req,group,leadTo,use,deg,comp,order,hSubCo,llh_sat_sub,llh_sat,llh_sat_sub_aux,step,deg_pos,nAllPosCon,nActPosCon,possImpactsPos,scoreLoS,wBus,wHum,wTru,wFun,wTim,wBou,wLif,wCom,wDat,last,addFun,func_dec,formulas,sat_formulas,possImpactsNeg,identify_conflict},%
  keywords=[2]{\#const,\#show,\#minimize,\#maximize,\#base,\#theory,\#count,\#external,\#program,\#script,\#end,\#heuristic,\#edge,\#project,\#show},%
  keywords=[3]{&,&dom,&sum,&diff,&show,&minimize},%
  morecomment=[l]{\#\ },%
  morecomment=[l]{\%\ },%
  commentstyle={\color{darkgray}}%
}
\lstset{numbers=left,numberblanklines=false,basicstyle=\ttfamily\small,language=clingo}
\lstset{escapeinside={*@}{@*}}

% Title and Author Information Must Immediately Follow
% the pdfinfo within the preamble
%
\title{An Answer Set Programming Framework for Reasoning about Agents' Beliefs and Truthfulness of Statements} %\\
\author{Marcello Balduccini$^1$ \And 
Michael Gelfond$^2$ \And
Enrico Pontelli$^3$ \and Tran Cao Son$^3$ 
\affiliations
$^1$Saint Joseph's University \and 
$^2$Texas Tech University\and 
$^3$New Mexico State University
\emails
   mbalducc@sju.edu,   
   michael.gelfond@ttu.edu,
\{epontell, tson\}@cs.nmsu.edu 
} %\\

@@

Artificial Intelligence research, in the area of knowledge-based and intelligent agents, has been progressing at
a rapid pace, making it possible to develop agents that can assist and/or replace humans in 
several tasks. Organizations  use  web-bots for interacting with clients in various capacities, e.g., in providing information about the company or making offers. This trend continues to be fueled by the ubiquitous use of online resources.

Unfortunately, not every business on the Internet is as 
honest as one would hope. Stories about businesses that cheat people of goods or services, or falsely advertise their services are not uncommon. Indeed, lying and misrepresentations are more widespread in online negotiations than 
offline ones \cite{elie}. Dishonesty is also present in
the news media, leading to the dissemination of untruthful stories  (e.g., the `Phuc Dat Bich' story\footnote{ 
{\tiny
\url{http://gizmodo.com/phuc-dat-bich-is-a-massive-phucking-faker-1744588099}
}} reported by BBC and Sky News). Another example is the case of a hotel reviewer with the alias {\em ``AmishBoy''} who gave high ratings to different hotels in the same day \cite{MinnichCMLF15}---thus suggesting  that his statements may be untrustworthy.

The concern about misinformation, disinformation, or dishonesty is also reflected in the 
development of resources that allow people to rate companies (e.g., \url{expedia.com}, \url{yelp.com},  \emph{Angie's List}) or defend the reputation of a company or an entity (e.g., \url{reputation.com}) or check for the accuracy of some information disseminated by a person or a new outlet (e.g., \url{factcheck.org}). All these aim at providing users with tools for checking the trustworthiness of a piece of information.

The above development highlights the problem of  \emph{understanding whether statements made by an agent can be accepted as truthful with respect to a context composed of observations,knowledge about default behavior, and corresponding beliefs.} This problem is not new to human. It is indeed as old as the existence of human on the earth. Yet, the amount of information that a human is receiving everyday has gone up significantly since the earth becomes flat, with the introduction of the Internet, and the advances in mobile technologies. This requires, ideally, tools that automatically filter out untrustworthy information for an user. The development of such a tool, on the other hand, requires the understanding of the reasoning process that helps a human to conclude that a piece of information receiving from a third party is true or false.

% \memo{I modified this part - up to this point: dividing into paragraphs, adding the last one}

@@

Consider the following motivating example.
\begin{example} \label{ex1}
When we first met at an event (time $t_0$), John stated that he comes from a poor family. 
Later, we  learn that John attends a private college, renowned for  its expensive tuition (time $t_1$). Later yet, we learn 
that John attends the college thanks to a  scholarship awarded  due to his financial hardship (time $t_2$).

This story spans over three time instants: $t_0$, $t_1$, and $t_2$. During the story, we learn (through observations) some facts, which  affect our beliefs about the world as follows:

\begin{itemize}

\item Time $t_0$: John makes the statement that his family is poor (property $poor$ is true). No observations are available to us and thus we do not know if John's family is indeed poor or not. We have no reasons to conclude that John's statement is not truthful.

\item Time $t_1$:  We observe that John attends an expensive college (property $in\_college$ is true). {\emph{Normally}, we believe that someone attending an expensive college is from a rich family} (default $d_1$). Hence, our belief at this point is that John is rich. This prevents us from accepting John's statement as truthful. We indicate that  the default $d_1$ is the reason to draw such conclusion.

\item Time $t_2$: We observe that John has a need-based scholarship   (property $has\_scholarship$ is true). {It is our belief that a student's need-based scholarship is \emph{usually} derived from the family's financial situation} (default $d_2$). Because both defaults are applicable and in conflict with each other, we believe neither that John's family is rich, nor that it is poor. This allows us to withdraw our previous conclusion that John's statement could not be accepted as truthful.

\item The situation might be different if we had a preference between our defaults. For instance, if we were inclined to favor $d_2$ over $d_1$, then we would believe that John's family is poor. This may strengthen our conclusion that John's statement can be accepted as truthful. 
% 
\end{itemize}  
\end{example}

The main contributions of this paper can be summarized as follows:
\begin{enumerate}
\item The formalization of an abstract model to 
represent and reason about the evolution of an agent's beliefs over time and to draw conclusions about whether third-party statements can be accepted as truthful;   
\item A concrete realization of the model using
Answer Set Programming; and
\item A demonstration of the proposed framework on an important problem in the
        area of  cyber security. 
\end{enumerate}
%The paper illustrates  how, starting
% from ({\em (i)}) a collection of observations about the (possibly evolving) state of the
% world, {\em (ii)}
% observations about (some of) the  actions performed by the agent, and 
%{\em (iii)}
% the observer's own bias about the \emph{normal} behavior of agents,
% an observer can draw conclusions about the truthfulness of statements made by
% the observed agent. 
 We illustrate the framework 
 using examples and discuss possible extensions that need to be considered. 
%We also discuss a potential application of the proposed framework.

@@

The language of $\cal B$ is based on a signature ${\cal K} = ({\cal L}, {\cal A})$, where
$\cal L$ is a set of ground atoms, referred to as \emph{fluents},
 and $\cal A$ is a set of \emph{action names}. A fluent literal is a literal
 built on $\cal L$.
 A domain description of $\cal L$ is composed of a collection of \emph{axioms} of 
 the following types:
\begin{itemize} 
 \item \emph{Executability Axioms:} these describe the conditions that enable the
         execution of an action, and they have the form $a \executable \ell_1\wedge \dots\wedge \ell_n$,
         where $a\in {\cal A}$ and $\ell_1,\dots,\ell_n$ are fluent literals.
  \item \emph{Dynamic Causal Laws:} these descriptions the direct effects of the execution
          of an action, and they have the form
          $a \causesif{\ell} \ell_1\wedge \dots \wedge \ell_n$
          where $a\in {\cal A}$ and $\ell,\ell_1,\dots,\ell_n$ are fluent literals.
  \item \emph{Static Causal Laws:} these description static dependencies among fluents, and they
          have the form
          $\ell \ifthen \ell_1\wedge\dots\wedge\ell_n$
          where $\ell, \ell_1,\dots,\ell_n$ are fluent literals.
 \end{itemize} 
 A planning problem in $\cal K$ is a tuple $(D, I, O)$, where $D$ is a domain description
 of $\cal K$ and $I, O$ are two consistent sets of fluent literals---denoting, respectively,
 the initial state and the goals of the planning problem.

@@

In this section, we propose a general framework for representing, and reasoning about, the beliefs held by an agent. The framework also establishes a link between such beliefs and whether the agent may accept as truthful statements made by a third party. The
framework  can be instantiated using specific paradigms for reasoning about actions and change and for non-monotonic reasoning. We build upon the following foundations:
%
 \begin{itemize}
%
\item An \emph{action signature} $\langle F,B,A \rangle$ is given, where $F$ are symbols for fluents (properties of the world whose truth value may change over time),
        $B$ are symbols for an agent's beliefs and $A$ are symbols for actions. The components of the signatures are pairwise disjoint.\footnote{This requirement is imposed to simplify some aspects of the presentation, but is not essential.} The notions of fluent literals and of set $\Sigma$ of states of the domain are introduced as usual.
\item It is possible to  observe the properties of the world and the occurrences of the actions    over time
        (e.g., we observe that John buys a car, John is a student, etc.). 
        Let $H^n$ be the set of all observations up to step $n$ and all recorded action occurrences up to step $n-1$.

\item We have adequate knowledge about actions  and their effects (e.g., the action of buying a car requires that the agent has money and its execution will result in the agent owning a car). This knowledge is represented by an action theory $Act$ in a suitable logic that allows reasoning about actions' effects and consequent changes to the world. 
        Let $\Phi_{Act}: \Sigma \times A \rightarrow 2^\Sigma$ denote a suitable transition function describing the evolution of the dynamic system of interest. $\hat\Phi_{Act}$ denotes the natural extension of $\Phi_{Act}$ to a history. %A resulting sequence $<\sigma_0,a_0,\sigma_1,a_1,\ldots,\sigma_k>$ is called a \emph{possible path}. 
        %Let us denote with $\models_A$ the entailment relation defined within the logical framework $A$ used to describe $Act$.

@@

The set of observations $O_w\subseteq H^n$ in Example~\ref{ex1} includes the observations such as 
%`John comes from a poor family' at time point $t_0$, 
\begin{itemize}
    \item `John attends an expensive college' at time point $t_1$, and 
    \item `John receives a need-based scholarship' at time point $t_2$.
\end{itemize}
 In this particular example we do not have any  action occurrences, i.e., $O_a = \emptyset$.\footnote{Some frameworks for reasoning about actions and change may require the introduction of a NOP action for modeling the example.} Our default theory $Def$ consists of defaults $d_1$ and $d_2$ stated earlier, which allow us to make conclusions regarding our agent's beliefs.

\iffalse
%TEMP
\begin{figwindow}[1,r,%
        {\includegraphics[width=0.48\textwidth]{newstep.pdf}},%
        {Answering Question: $(H^n,Act,Def) \models p[t]$?\label{fig:schema} }]
\end{figwindow}
\fi
Let us consider a theory $T = (H^n, Act, \textit{Def})$ and the associated $\Phi_{Act}$ and $\models_D$. First, we are interested in defining the beliefs held by the agent over time. And, then, we are interested in using this information for answering the question of whether a statement $b[t]$,  made at time step $t$, where $b \in B$, can be accepted as truthful. For this, we define the entailment relation $\models$ between $T$ and $b[t]$. Intuitively, the process entails the steps below: %(Figure~\ref{fig:schema}):

\begin{itemize}
\item Compute  $\hat\Phi_{Act}(H^n,\Sigma_0)$, the set of states that are reachable from the initial state given the history $H^n$,
where $\Sigma_0$ denotes the initial state specified by $Act$ and $H^n$, which can be incomplete. 
For compactness, below we abbreviate $\hat\Phi_{Act}(H^n,\Sigma_0)$ with $W[n]$;

%Compute the set $W[t]$ of fluent literals that are expected to hold at time step \(t\) according to the history;

\iffalse
\memo{
Compute  $\hat\Phi_{Act}(H^n,\Sigma_0)$, the set of states that are reachable from the initial state given the history $H^n$,
where $\Sigma_0$ denotes the initial state specified by $Act$ and $H^n$. 
For simplicity of the presentation, we denote $\hat\Phi_{Act}(H^n,\Sigma_0)$ with $W[t]$. 
}
\fi

\item Determine  the agent's beliefs by finding which propositions are derivable from $\textit{Def}$ and $W[t]$ (using $\models_D$). We note that $W[t]$ can be a set of states and the different notions of entailment such as \emph{skeptical} and \emph{credulous} entailment could be applied here;

\item Determine if the statement $b[t]$ is consistent with respect to the agent's beliefs. 
\end{itemize}  
%
%The definition of ``expected to hold'' depends on the choice of logic and approach. 
%For the purpose of our discussion, $W[t]$ denotes the set $\hat\Phi_{Act}(H^n)$ of states that are reachable from the initial state given the history $H^n$.
%, where $\Sigma_0$ denotes the initial state specified by $Act$ and $H^n$. 
%For simplicity of the presentation, we denote $\hat\Phi_{Act}(H^n,\Sigma_0)$ with $W[t]$. 
%we assume that $W[t] = \{ l \,|\, l\in \bigcap \hat\Phi_{Act}(H^t) \}$ characterizes
%the fluent literals that are expected to hold at time step $t$ according to history $H^n$ given $Act$. 
%
The entailment relation between $T$ and $b[t]$ can thus be defined by:
%
%T \models p[t] \Leftrightarrow
%\begin{array}{c} 
%%T \models p[t] \\
%% \textit{if and only if} \\
%  \forall W[t]. \left(\begin{array}{c}
%   W[t] = \{z \mid Act \cup O_a \cup O_w \models_A z[t]\} \Rightarrow \\
%                                                                 Def \cup W[t] \models_D p
%                                                                \end{array}\right)
%\end{array} 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{align} 
T \models b[t] \,\,\Leftrightarrow\,\, \langle W[t], Def \rangle \models_D b \label{entail} 
\end{align}

Note that this definition also allows one to identify elements of $H^t$ that, when obtained, will result in the 
confirmation or denial of $T \models b[t]$. As such, a system that obeys \eqref{entail} can also be used by users who are 
interested in  what they need to do in order to believe in a statement about $b$ at the time step $t$,
 given their beliefs about the behavior of the observed agents.

%
%\smartdiagram[flow diagram:horizontal]{Edit,
%  \LaTeX, Bib\TeX/ biber, make\-index, \LaTeX}
\iffalse\begin{figure}
\centering
\resizebox{1.6in}{!}{% 
\smartdiagram[priority descriptive diagram]{
  $W[t]$ model of $A \cup O_a \cup O_w$ restricted to time step $t$,
  $\forall W[t]. (D \cup W[t] \models_D p)$?}
}  
\caption{Answering $(O_a,O_w,A,D) \models p[t]$?}
\label{fig:schema} 
\end{figure} \fi
%
To develop a concrete system for reasoning about truthfulness of agents using \eqref{entail}, specific formalizations of $Act$ and $\textit{Def}$ need to be developed. There is a large body of research related to these two areas, and deciding which one to use depends on the system developer. 
Well-known formalisms for reasoning about actions and change, such as action languages \cite{GelfondL98}, situation calculus \cite{Reiter01}, etc., can be employed for $Act$ and $\Phi_{Act}$.  
Approaches to default reasoning with preferences, such as those proposed in  \cite{brew99,BrewkaE00,DelgrandeST03,GelfondS98}), can be used for $\textit{Def}$ (and $\models_D$). In addition, 
let us note that, in the literature, $\models_D$ can represent \emph{skeptical} or \emph{credulous} reasoners; and  the model does not specify how observations are collected. Deciding which type of reasoning is suitable or how to collect observations is an important issue, but it is application-dependent and beyond  the scope of this paper. 
%In this paper, we use observations  in a broad sense: it might be something that we learn some facts about a property of the world, or we actually observed it (via the execution of an action), etc. 
%

\section{Reasoning about Beliefs and Truthfulness Using Answer Set Programming}

We will now present a concrete system for reasoning about truthfulness of agents using \emph{Answer Set Programming (ASP)} \cite{MarekT99,GelfondL91}.  We select ASP as the host language since there exist several approaches to reasoning about actions and changes as well as default reasoning using ASP. This enables  the seamless integration of the two steps involved in ~\eqref{entail} into a single system. 
Several efficient ASP solvers are available, that enable the use of the proposed system in practical applications. In particular, we will use answer sets (and projections of answer sets over specific time points) to capture $W[t]$, and use established encodings of action theories and defaults in ASP.

@@

A logic program $\Pi$ is a set of rules of the form
%
\begin{equation}
\label{lprule1}
 a_0 \leftarrow a_1,\ldots,a_m,\naf a_{m+1},\ldots,\naf a_n
\end{equation}
%
or
%
\begin{equation}
\label{lprule2}
 \bot \leftarrow a_1,\ldots,a_m,\naf a_{m+1},\ldots,\naf a_n
\end{equation}
%
where $0 \le m \le n$, each $a_i$ is an atom of a first-order
language $\cal P$, $\bot$ is a special symbol denoting the
truth value \emph{false}, and $\naf$ is a connective called {\em
negation-as-failure}. A negation as failure literal (or
naf-literal) is of the form $\naf a$ where $a$ is an atom. For a
rule of the form (\ref{lprule1})-(\ref{lprule2}), the left and
right hand side of the rule are called the \emph{head} and the
\emph{body}, respectively. A rule of the form (\ref{lprule2}) is
also called a constraint.

Given a logic program $\Pi$, we will assume that each rule in
$\Pi$ is replaced by the set of its ground instances so that all
atoms in $\Pi$ are ground. Consider a set of ground atoms $X$. The
body of a rule of the form (\ref{lprule1}) or (\ref{lprule2}) is
satisfied by $X$ if $\{a_{m+1},\ldots,a_n\} \cap X = \emptyset$
and $\{a_{1},\ldots,a_m\} \subseteq X$. A rule of the form
(\ref{lprule1}) is satisfied by $X$ if either its body is not
satisfied by $X$ or $a_0 \in X$. A rule of the form
(\ref{lprule2}) is satisfied by $X$ if its body is not satisfied
by $X$. An atom $a$ is supported by $X$ if $a$ is the head of some
rule of the form (\ref{lprule1}) whose body is satisfied by $X$.

For a set of ground atoms $S$ and a program $\Pi$, the reduct
of $\Pi$ with respect to $S$, denoted by $\Pi^S$, is the program
obtained from the set of all ground instances of $\Pi$ by deleting

\begin{enumerate}
\item each rule that has a naf-literal $\naf a$ in its body with
$a \in S$, and

\item all naf-literals in the bodies of the remaining clauses.
\end{enumerate}
